CLEM-ICD - Clinical Language Encoding with ModernBERT \\\large{Using ModernBERT to Improve Automated ICD-10 Classification}

### 2025-03-23 article/article.qmd
cleanup

---
title: CLEM-ICD - Clinical Language Encoding with ModernBERT
subtitle: Using ModernBERT to Improve Automated ICD-10 Classification
author:
  - name: Tyler Cross
    email: tyler.cross@berkeley.com
    affiliations: 
        - id: some-u
          name: University of California, Berkeley
          department: School of Information
          address: 102 South Hall
          city: Berkeley
          state: CA
          postal-code: 94720
    attributes:
        corresponding: true
    note: |
      Final project submission for the UC Berkeley Master of Information and Data Science program, DATASCI 266: Natural Language Processing with Deep Learning
abstract: |
  Medical coding is a critical yet complex process that requires translating clinical documentation into standardized ICD-10 codes. Currently, healthcare providers rely on dedicated medical coders who manually review extensive clinical notes to assign appropriate codes from 150,000+ possible diagnostic and procedural codes. This project builds upon the work of @edin2024explainable with two key innovations: (1) replacing RoBERTa with ModernBERT [@warner2024modernbert] as the encoder to potentially improve performance through its enhanced context window, and (2) adding a LLM-powered reasoning layer that provides natural language explanations for code predictions, bridging the gap between model predictions and human validation. Using the MIMIC-IV dataset, we implement and evaluate this approach using standard metrics (precision, recall, F1) and explanation quality.
keywords: 
  - ModernBERT
  - natural language processing
  - MIMIC-IV
  - ICD-10
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

>>>>

---
title: CLEM-ICD - Clinical Language Encoding with ModernBERT
subtitle: Using ModernBERT to Improve Automated ICD-10 Classification
author:
  - name: Tyler Cross
    email: tyler.cross@berkeley.com
    affiliations: 
        - id: some-u
          name: University of California, Berkeley
          department: School of Information
          address: 102 South Hall
          city: Berkeley
          state: CA
          postal-code: 94720
    attributes:
        corresponding: true
    note: |
      Final project submission for the UC Berkeley Master of Information and Data Science program, DATASCI 266: Natural Language Processing with Deep Learning
abstract: |
  Medical coding—the assignment of standardized ICD-10 codes to clinical documentation—remains a labor-intensive process requiring expert manual review of extensive narratives against 150,000+ potential classifications. We present CLEM-ICD, an extension of the PLM-ICD framework that replaces RoBERTa with ModernBERT to leverage its 8192-token context window for improved automated medical coding. Our architectural enhancement preserves the multi-label classification paradigm while significantly expanding the model's capacity to process lengthy clinical text, capturing distant dependencies crucial for accurate code assignment. Experiments on the MIMIC-IV dataset demonstrate that CLEM-ICD achieves superior performance compared to previous approaches, as measured by precision, recall, and F1-score. This work addresses a critical bottleneck in healthcare administration through advanced NLP techniques, offering a scalable solution for reducing the cognitive burden of medical coding while maintaining diagnostic accuracy.
keywords: 
  - clinical natural language processing
  - medical coding automation
  - transformer models
  - multi-label classification
  - long-context language models
  - ModernBERT
  - MIMIC-IV
  - ICD-10
  - healthcare informatics
  - biomedical text classification
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

### 2025-03-23 article/article.tex
cleanup

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler Cross},
  pdfkeywords={ModernBERT, natural language
processing, MIMIC-IV, ICD-10},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Medical coding is a critical yet complex process that requires
translating clinical documentation into standardized ICD-10 codes.
Currently, healthcare providers rely on dedicated medical coders who
manually review extensive clinical notes to assign appropriate codes
from 150,000+ possible diagnostic and procedural codes. This project
builds upon the work of \citet{edin2024explainable}{]} with two key
innovations: (1) replacing RoBERTa with ModernBERT
\citep{warner2024modernbert} as the encoder to potentially improve
performance through its enhanced context window, and (2) adding a
LLM-powered reasoning layer that provides natural language explanations
for code predictions, bridging the gap between model predictions and
human validation. Using the MIMIC-IV dataset, we implement and evaluate
this approach using standard metrics (precision, recall, F1) and
explanation quality.
\end{abstract}

\begin{keyword}
    ModernBERT \sep natural language processing \sep MIMIC-IV \sep 
    ICD-10
\end{keyword}
\end{frontmatter}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CLEM-ICD - Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler Cross},
  pdfkeywords={clinical natural language processing, medical coding
automation, transformer models, multi-label classification, long-context
language models, ModernBERT, MIMIC-IV, ICD-10, healthcare
informatics, biomedical text classification},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Medical coding---the assignment of standardized ICD-10 codes to clinical
documentation---remains a labor-intensive process requiring expert
manual review of extensive narratives against 150,000+ potential
classifications. We present CLEM-ICD, an extension of the PLM-ICD
framework that replaces RoBERTa with ModernBERT to leverage its
8192-token context window for improved automated medical coding. Our
architectural enhancement preserves the multi-label classification
paradigm while significantly expanding the model's capacity to process
lengthy clinical text, capturing distant dependencies crucial for
accurate code assignment. Experiments on the MIMIC-IV dataset
demonstrate that CLEM-ICD achieves superior performance compared to
previous approaches, as measured by precision, recall, and F1-score.
This work addresses a critical bottleneck in healthcare administration
through advanced NLP techniques, offering a scalable solution for
reducing the cognitive burden of medical coding while maintaining
diagnostic accuracy.
\end{abstract}

\begin{keyword}
    clinical natural language processing \sep medical coding
automation \sep transformer models \sep multi-label
classification \sep long-context language
models \sep ModernBERT \sep MIMIC-IV \sep ICD-10 \sep healthcare
informatics \sep 
    biomedical text classification
\end{keyword}
\end{frontmatter}
